---
title: "Mapping from CDASH to SDTM Using R"
author: "Instructor Name"
date: "`r Sys.Date()`"
format: html
---

# Walkthrough: Mapping from CDASH to SDTM Using R

## Objective

The goal of this walkthrough is to map clinical trial data from a CDASH-compliant dataset to an SDTM dataset using R. We will use a mock dataset that simulates the CDISC CDASH Demographics CRF and apply tidyverse packages to transform the data according to CDISC standards.

## Step 1: Load Required Libraries

Load the required packages for this exercise, tidyverse for data manipulation.

```{r}
# Install and load necessary packages
# 'tidyverse' is a collection of popular data manipulation packages that help streamline the data transformation process.

if (!requireNamespace("tidyverse", quietly = TRUE)) {
  install.packages("tidyverse")
}

library(tidyverse)    # Collection of R packages for data science
```

## Step 2: Generate the Mock CDASH Dataset

We will generate a mock dataset based on the CDISC CDASH Demographics CRF, which includes key variables such as BRTHDAT, AGE, AGEU, SEX, ETHNIC, RACE, and RACEOTH.

```{r}
# Generating Mock CDASH Demographics Dataset
set.seed(123)  # Set seed for reproducibility

# Create mock data
cdash_data <- data.frame(
  STUDYID = rep("CDASH_DEMO_01", 20),
  SUBJID = sprintf("SUB%03d", 1:20),
  BRTHDAT = sample(seq(as.Date('1940/01/01'), as.Date('2005/12/31'), by="day"), 20),
  AGE = sample(18:85, 20, replace = TRUE),
  AGEU = rep("Years", 20),
  SEX = sample(c("F", "M", "U", "UNDIFFERENTIATED"), 20, replace = TRUE),
  ETHNIC = sample(c("HISPANIC OR LATINO", "NOT HISPANIC OR LATINO", "NOT REPORTED", "UNKNOWN"), 20, replace = TRUE),
  RACE = sample(c("AMERICAN INDIAN OR ALASKA NATIVE", "ASIAN", "BLACK OR AFRICAN AMERICAN", "NATIVE HAWAIIAN OR OTHER PACIFIC ISLANDER", "WHITE", "NOT REPORTED", "UNKNOWN", "OTHER"), 20, replace = TRUE),
  RACEOTH = ifelse(runif(20) < 0.1, sample(c("Caucasian", "Black American"), 20, replace = TRUE), NA)  # Some entries have other race specified
)

# Display the first few rows of the mock dataset
head(cdash_data)
```

## Step 3: Understanding CDASH Data

In this dataset, we have demographic information collected at the point of data acquisition (e.g., SUBJID, AGE, SEX, RACE). These fields are collected using CDASH standards to ensure consistency and ease of mapping.

## Step 4: Map CDASH Data to SDTM DM Domain

Now we will map the data from the CDASH dataset to an SDTM DM (Demographics) domain dataset using the dplyr transformations.

```{r}
# Map CDASH data to SDTM DM domain
sdtm_dm <- cdash_data %>%
  mutate(
    DOMAIN = "DM",             # SDTM Domain Identifier
    USUBJID = paste0(STUDYID, "-", SUBJID), # Unique Subject Identifier
    RACERECOD = ifelse(RACE == "OTHER" | RACE == "UNKNOWN",
                       ifelse(!is.na(RACEOTH) & RACEOTH == "Caucasian", "WHITE",
                              ifelse(!is.na(RACEOTH) & RACEOTH == "Black American", "BLACK OR AFRICAN AMERICAN", RACE)),
                       RACE) # Retain RACE value if RACEOTH is NA
  ) %>%
  transmute(
    STUDYID = STUDYID,          # Study Identifier
    DOMAIN = DOMAIN,            # Domain Identifier
    USUBJID = USUBJID,          # Unique Subject Identifier
    SUBJID = SUBJID,            # Subject Identifier for the Study
    BRTHDAT = BRTHDAT,          # Birth Date
    AGE = AGE,                  # Age of the Subject
    AGEU = AGEU,                # Unit for Age
    SEX = SEX,                  # Sex of the Subject
    ETHNIC = ETHNIC,            # Ethnicity of the Subject
    RACE = RACE,                # Original Race variable
    RACERECOD = RACERECOD,      # Recoded Race variable
    RACEOTH = RACEOTH           # Original RACEOTH variable
  )

# View the mapped SDTM DM dataset
head(sdtm_dm) # View the first few rows to validate the mapping
```

## Step 5: Review the Mapping Process

-   **STUDYID**: Study identifier, derived from STUDYID in the CDASH dataset.
-   **DOMAIN**: A constant value "DM" to indicate the Demographics domain.
-   **USUBJID**: Unique subject identifier, created by concatenating STUDYID and SUBJID.
-   **BRTHDAT**: Birth date of the subject.
-   **AGEU**: Unit for age, taken directly from AGEU in the dataset.
-   **SEX / ETHNIC / RACE**: Subject demographic information, mapped directly from CDASH.
-   **RACERECOD**: A derived variable combining RACE and RACEOTH where applicable, without altering the original RACE or RACEOTH. Specifically, "Caucasian" is mapped to "WHITE" and "Black American" is mapped to "BLACK OR AFRICAN AMERICAN".

## Step 6: Create Define-XML Metadata

To document the SDTM dataset and provide context for regulatory submissions, we use metadata. Below is a simplified approach for generating metadata using R.

```{r}
# Example metadata definition (for demonstration purposes)
# This metadata documents each variable in the SDTM DM dataset, providing descriptions and types.
define_metadata <- data.frame(
  variable = c("STUDYID", "DOMAIN", "USUBJID", "SUBJID", "BRTHDAT", "AGE", "AGEU", "SEX", "ETHNIC", "RACE", "RACEOTH", "RACERECOD"),
  label = c(
    "Study Identifier",
    "Domain Abbreviation",
    "Unique Subject Identifier",
    "Subject Identifier for the Study",
    "Birth Date of the Subject",
    "Age of the Subject",
    "Age Unit",
    "Sex of the Subject",
    "Ethnicity of the Subject",
    "Race of the Subject",
    "Other Race (if applicable)",
    "Recoded Race including corrected RACEOTH"
  ),
  type = c("Char", "Char", "Char", "Char", "Date", "Num", "Char", "Char", "Char", "Char", "Char", "Char")
)

# View metadata
print(define_metadata)
```

## Summary

By the end of this walkthrough, participants should understand: - The relationship between **CDASH and SDTM** standards. - How to map data from a CDASH-compliant dataset to an SDTM dataset. - The importance of documenting the dataset structure using metadata for regulatory compliance. - How to iteratively clean and correct data to ensure compliance.

This hands-on approach will reinforce the importance of standardization in clinical trial data management and prepare students for real-world data submission workflows.
